The state learner -- a super learner for right-censored data

### 2024-05-28 tex/arxiv-version/manuscript.tex
arxiv link

\documentclass[12pt]{article}
\usepackage[a4paper,left=2.8cm, right=2.8cm, top=3cm, bottom=3cm]{geometry}

@@

\theoremstyle{thmstyleone}%\
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

%% Line spacing
\linespread{1.65}

% \usepackage{lineno}
% \linenumbers

%% Only for comments:
\usepackage[author=]{fixme}
\fxusetheme{color}
\definecolor{fxtarget}{rgb}{.5,.5,.5}
\definecolor{fxnote}{rgb}{.5,.5,.5}
\fxsetup{status=draft}

@@

%% Title
\title{\Large The state learner -- a super learner for right-censored data}
\date{\small \today}

\usepackage{authblk}
\author[1,*]{\normalsize Anders Munch}
\author[1]{\normalsize Thomas~A.~Gerds}
\affil[1]{\normalsize Section of Biostatistics, University of Copenhagen}
\affil[*]{\normalsize Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}

@@

% \keywords{Competing risks, cross-validation, loss based estimation,
% right-censored data, super learner}

@@

\pagebreak

@@

% Medical decision making is traditionally based on the discrepancy
% between intuitive judgment and the law of chance
% \citep{redelmeier1995probability}.
% We are concerned with the statistical models that provide
% probabilistic predictions about the medical future of a patient. We
% have two different applications of these models in mind. One is as a
% stand-alone tool which predicts the likelihood of medical outcomes.
% The other is as nuisance parameter estimate
% input to an algorithm which estimates a causal target
% parameter such as an average treatment effect and depends on estimates
% of nuisance parameters such as the conditional probability of being
% uncensored.  An example of the latter is the targeted minimum loss
% based estimator \citep{rytgaard2022targeted}. In both applications,
% the models have to be learned based on available data from previous
% patients.
A super learner is a machine learning algorithm that combines a finite
set of learners into a meta learner by estimating prediction
performance in hold-out samples using a pre-specified loss function
\citep{van2007super}. When the aim is to make a prediction model,
super learners combine strong learners, such as Cox regression models
and random survival forests \citep[][Section 8.4]{gerds2021medical}.
% For targeted learning of a low dimensional
% target parameter in presence of a high-dimensional nuisance parameter,
% a super learner will usually also include the highly adaptive lasso
% \citep{van2011targeted, benkeser2016highly,van2017generally}.
While the general idea of combining strong learners based on cross-validation
stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
name super learner is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}.
% The super learner is guaranteed to perform
% almost as well as the model which minimises the expected performance, i.e., the
% model we would select if we could evaluate the prediction performance in an
% infinite hold-out sample.

@@

Machine learning based on right-censored data commonly uses the
partial log-likelihood as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
However, this loss function does not work well with data splitting
(cross-validation) because the partial log-likelihood loss assigns an
infinite value when applied to a test set time point which does not
occur in the learning set as soon as the learner predicts piece-wise
constant cumulative hazard functions. This is the case for prominent
survival learners including the Kaplan-Meier estimator, the random
survival forest, and the semi-parametric Cox regression model.  When a
proportional hazards model is assumed, the baseline hazard function
can be profiled out of the likelihood \citep{cox1972regression}.  The
cross-validated partial log-likelihood loss \citep{verweij1993cross}
has therefore been suggested as a loss function for super learning
which however restricts the library of learners to include only Cox
proportional hazards models \citep{golmakani2020super}.
% A lesser
% known fact is that a super learner constructed with the partial
% log-likelihood loss implicitly depends on the censoring distribution
% \citep{hjort1992inference,whitney2019comment}.

Alternative approaches for super learning with right-censored data use
an inverse probability of censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
% An IPCW loss function is attractive because the associated risk does
% not depend on the censoring distribution but describes a feature of
% the population of interested. Similar results can be obtained using
censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
All these methods rely on an estimator of the censoring distribution,
and their drawback is that this estimator has to be pre-specified.
% When the data-generating mechanism is complex and not
% well-understood, pre-specification of the censoring distribution is a
% challenge.
% Inverse probability of censoring weighted loss has been studied for
% decades \citep{graf1999assessment, hothorn2006survival,
  % gerds2006consistent}. A disadvantage of inverse probability of
% censoring weighted loss estimators is that they require a
% pre-specified model for the censoring distribution.
% Recently \cite{westling2021inference} proposed a procedure which
% iterates between super learning of the censoring distribution and
% super learning of the event time distribution.
An approach which avoids a pre-specified censoring model was proposed
independently by \cite{han2021inverse} and
\cite{westling2021inference}. In both articles, the authors suggest to
iterate between learning of the outcome model and learning of the
censoring model using IPCW loss functions. However, no general
theoretical guarantees seem to exist for this procedure, and it has
not yet been extended to the situation with competing risks.

@@

In a competing risk framework \citep{andersen2012statistical}, let
\( T\) be a time to event variable, \(D\in\{1,2\}\) the cause of the
event, and $X \in \mathcal{X}$ a vector of baseline covariates taking
values in a bounded subset \( \mathcal{X} \subset \R^p \),
\( p\in\N \). Let $\tau< \infty$ be the prediction horizon. We use
\( \mathcal{Q} \) to denote the collection of all probability measures
on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \) such that
\( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard
functions are defined by
\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \)
such that
\begin{equation*}
  % \label{eq:cum-haz}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*}
For ease of presentation we assume throughout that the map
\( t\mapsto \Lambda_j(t \mid x) \) is continuous for all \( x \) and
\( j \). This is not a limitation: All arguments carry over directly
to the general case. We denote by \(S\) the conditional event-free
survival function:
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterised by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)=& \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=1\}}}\\
                         &  \left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}_{\tau}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

@@

As input to the super learner we need a data set
\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in \mathcal{P} \) and a collection of candidate
learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}_{\tau}\) which takes a data
set as input and returns an estimate $a(\data_n) \in \mathcal{M}_{\tau}$ of
$\Lambda_{j}$.
% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
In what follows, we use the short-hand notation
\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
performance of \(a \in \mathcal{A}\) with a loss function
\(L\colon \mathcal{M}_{\tau} \times \mathcal{O} \rightarrow \R_+\) and
estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
cross-validation. Specifically, the expected loss of $a\in\mathcal A$
is estimated by splitting the data set $\data_n$ into $K$ disjoint
approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  % \label{eq:cv-risk-est}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
% The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
% learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
The oracle learner is defined as the learner that minimises the
expected loss under the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner
depend on the library of learners and on the number of folds \(K\),
and that the oracle learner is a function of the data and the unknown
data-generating distribution. However, these dependencies are
suppressed in the notation.

@@

% The problem with most existing super learners for right-censored data is that
% they depend on a pre-specified estimator of the censoring distribution.

@@

% (Figure \ref{fig:multi-state-process}).
% \begin{figure}[h]
  % \centering %
  % \includeFigCond[width=.5\textwidth]{figure-multi-state-process.pdf}
  % \caption{Illustration of the multi-state process \(\eta\) used by
    % the state learner. Note that `censored' is a state, hence the
    % process is always observed at any time.}
  % \label{fig:multi-state-process}
% \end{figure}

@@

% \begin{equation}\label{eq:transition}
%   \begin{split}
%   F(t, 1, x)
%     &
%       % = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
%     = \int_0^t \exp{{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }}  \Lambda_{1}(\diff s \mid x),
%   \\
%   F(t, 2, x)
%     &
%       % = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
%       = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Lambda_{2}(\diff s \mid x),
%     \\
%     F(t, -1, x)
%     &
%       % = P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
%       = \int_0^t \exp{\{-\Lambda_{1}(s \mid x)-\Lambda_{2}(s \mid x) - \Gamma(s \mid x)\} }  \Gamma(\diff s \mid x),
%     \\
%     F(t, 0, x)
%     &
%       = P(\tilde{T} > t \mid X= x)
%       = 1- F(t, 1, x) - F(t, 2, x)- F(t, -1, x).
%   \end{split}
% \end{equation}

@@

censoring distribution. % We further
% define \(\mathcal{H}\) as the set of all probability distributions on
% \( \mathcal{X} \) and
% \begin{equation*}
%   \mathcal{H}_n=\left\{h\colon\mathcal{O}^n\longrightarrow\mathcal{H} \midd
%     h(\data_n)=H_n = \frac 1 n \sum_{i=1}^n \delta_{X_i} \right\}
% \end{equation*}
% as the library of learners which consists of a single learner, the empirical
% distribution function.

@@

% \begin{align*}
%   \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})
%   &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
%     \intertext{where in correspondence with  the relations in equation \eqref{eq:transition},} 
%     \phi_{a_1,a_2, b}(\data_n)(t,1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_1(\data_n)(\diff s \mid x),\\
%   \phi_{a_1,a_2, b}(\data_n)(t,2,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  a_2(\data_n)(\diff s \mid x),\\
%   \phi_{a_1,a_2, b}(\data_n)(t,-1,x) &= \int_0^t \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid x)\} }  b(\data_n)(\diff s \mid x),\\
%   \phi_{a_1,a_2, b}(\data_n)(t,0,x) &= 1
%                                       - \phi_{a_1,a_2, b}(\data_n)(t,1,x)
%                                       - \phi_{a_1,a_2, b}(\data_n)(t,2,x)
%     -\phi_{a_1,a_2, b}(\data_n)(t,-1,x)
% \end{align*}

@@

% We let \( \E_P \) denote
% expectation under \( P \) and use \( \Theta \) to denote the
% collection of all conditional state-occupation probability functions.
Proposition~\ref{prop:stric-prop} can be derived from the fact that the
integrated Brier score (also called the continuous ranked probability score) is
a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
if we minimise the average loss of the integrated Brier score, we recover the
parameters of the data-generating distribution. Specifically, this implies that
the oracle of a state learner is consistent if the library of learners contains
at least one learner that is consistent for estimation of \( F \). Recall that
the function \(F\) implicitly depends on the data-generating probability measure
\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
this dependence explicit by writing \(F_0\) for the function which is obtained
by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
\eqref{eq:transition}. In the following we let
\( \mathcal{H}_{\mathcal{P}} = \{F_P : P \in \mathcal{P}\} \) where \( F_P \) is defined as in
equation~(\ref{eq:F-def}) using the measure \( P \in \mathcal{P} \).

\begin{proposition}
  \label{prop:stric-prop}
  If \(P_0\in\mathcal{P}\) then
  \begin{equation*}
    F_0 = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]},
  % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
  \end{equation*}
  \( H \)-almost surely for any \( j\in \{-1,0,1,2\} \) and almost any
  \( t \in [0, \tau]\).
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:proof-proposition}.
\end{proof}

@@

We establish a finite sample oracle result for the state learner. Our
Corollary~\ref{cor:oracle-prop} is in essence a special case of
Theorem 2.3 in \citep{vaart2006oracle}.  We assume that we split the
data into equally sized folds, and for simplicity of presentation we
take \( n \) to be such that \( |\data_n^{-k}| = n/K \) with \( K \)
fixed. We will allow the number of learners to grow with \( n \) and
write
\( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
\mathcal{B}_n)\) as short-hand notation and to emphasise the
dependence on \( n \).
% The discrete super $\hat{\phi}_n$ and the
% oracle $\tilde{\phi}_n$ are defined as in Section~\ref{sec:framework}
% but now with respect to the library \( \mathcal{F}_n \) and the loss
% \( B \).
In the following we let the space \( \mathcal{H}_{\mathcal{P}} \) be equipped with the norm
\( \| \blank \|_{P_0} \) defined as
\begin{equation}
  \label{eq:norm}
  \| F \|_{P_0} = 
  \left\{
    \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
  \right\}^{1/2}.
\end{equation}

@@

The asymptotic distribution of \( \hat{\Psi}^0 \) is
difficult to analyse due to the cross-validated model selection step
involved in the estimation of the nuisance parameters $\Lambda_1$ and
$\Lambda_2$.
% In addition, the estimator will typically have an
% asymptotic bias that vanishes at a too slow 
% rate.
Using tools from semi-parametric efficiency theory, it is possible to
construct a so-called targeted or debiased estimator with an
asymptotic distribution which we know how to estimate
\citep{bickel1993efficient,van2011targeted,chernozhukov2018double}. A
targeted estimator is based on the efficient influence function for
the parameter $\tilde{\Psi}^0$ and relies on estimators of the
nuisance parameters $\Gamma$, \( \Lambda_1 \) and $\Lambda_2$. The
efficient influence function is a \( P \)-zero mean and square
integrable function which we denote by
\( \psi(\blank ; \Lambda_1, \Lambda_2, \Gamma) \). The name is
justified because any regular asymptotically linear estimator that has
\( \psi \) as its influence function is asymptotically efficient,
meaning that it has smallest asymptotic variance among all regular
asymptotically linear estimators \citep{bickel1993efficient}.

@@

% ANDERS: hedder det ikke Nelson-Aalen i stedet for Kaplan-Meier?
%         kan du skrive klart hvad outcome er i denne foerste del? er det combined endpoint som man kalder for tumor-free survival?  
For all super learners we use a library consisting of three learners:
The Kaplan-Meier estimator
\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model with
main effects \citep{cox1972regression, survival-package}, and a random
survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the
same library to learn the outcome distribution and the censoring
distribution. Note that the three learners in our library of learners
can be used to learn the cumulative hazard functions of the 
outcome and the censoring distribution. The latter works by
training the learner on the data set \( \data_n^c \), where
\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a
learner for the cumulative hazard function of the outcome to learn the
cumulative hazard function of the censoring time, we mean that the
learner is trained on \( \data_n^c \).

We compare the state learner to two IPCW based super learners: The
first super learner, called IPCW(Cox), uses a Cox model with main
effects to estimate the censoring probabilities, while the second
super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
estimate the censoring probabilities. The Cox model for the censoring
distribution is correctly specified in both simulation settings while
the Kaplan Meier estimator only estimates the censoring model
correctly in the simulation setting where censoring is
independent. Both IPCW super learners are fitted using the
\texttt{R}-package \texttt{riskRegression}
\citep{Gerds_Ohlendorff_Ozenne_2023}.
%
% | time | sim_setting | true_events | true_cens | at_risk |
% |------+-------------+-------------+-----------+---------|
% |   36 | original    |      24.619 |    61.853 |  25.774 |
% |   36 | indep_cens  |      24.674 |    38.740 |  46.141 |
%
The IPCW super learners use the integrated Brier score up to a fixed time
horizon (36 months). The marginal risk of the event before this time horizon is
\(\approx 24.6\)\%. Under the `dependent censoring' setting the marginal
censoring probability before the time horizon is \(\approx 61.9\)\%. Under the
`independent censoring' setting the marginal censoring probability before this
time horizon is \( \approx 38.7 \)\%.

@@

>>>>

\documentclass{article}
\usepackage[a4paper,margin=3cm]{geometry}

@@

\theoremstyle{plain}%\
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

% Handling new lines
\setlength{\parskip}{1em}
\setlength{\parindent}{0em}

% Handling space after sections
\usepackage{titlesec}
\titlespacing*{\section}{0em}{2em}{0em}
\titlespacing*{\subsection}{0em}{2em}{0em}
\titlespacing*{\subsubsection}{0em}{2em}{0em}

% No spacing after in start of list
\setlist[itemize]{topsep=0pt}
\setlist[enumerate]{topsep=0pt}

@@

%% Title
\title{The state learner -- a super learner for right-censored data}
\date{\today}

\usepackage{authblk}
\author[1,*]{Anders Munch}
\author[1]{Thomas~A.~Gerds}
\affil[1]{Section of Biostatistics, University of Copenhagen}

@@

@@

@@

A super learner is a machine learning algorithm that combines a finite
set of learners into a meta learner by estimating prediction
performance in hold-out samples using a pre-specified loss function
\citep{van2007super}. When the aim is to make a prediction model,
super learners combine strong learners, such as Cox regression models
and random survival forests \citep[][Section 8.4]{gerds2021medical}.
While the general idea of combining strong learners based on cross-validation
stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
name super learner is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}.

@@

Machine learning based on right-censored data commonly uses the
partial log-likelihood as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
However, this loss function does not work well with data splitting
(cross-validation) because the partial log-likelihood loss assigns an
infinite value when applied to a test set time point which does not
occur in the learning set as soon as the learner predicts piece-wise
constant cumulative hazard functions. This is the case for prominent
survival learners including the Kaplan-Meier estimator, the random
survival forest, and the semi-parametric Cox regression model.  When a
proportional hazards model is assumed, the baseline hazard function
can be profiled out of the likelihood \citep{cox1972regression}.  The
cross-validated partial log-likelihood loss \citep{verweij1993cross}
has therefore been suggested as a loss function for super learning
which however restricts the library of learners to include only Cox
proportional hazards models \citep{golmakani2020super}.

Alternative approaches for super learning with right-censored data use an
inverse probability of censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
methods rely on an estimator of the censoring distribution, and their drawback
is that this estimator has to be pre-specified. An approach which avoids a
pre-specified censoring model was proposed independently by
\cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
authors suggest to iterate between learning of the outcome model and learning of
the censoring model using IPCW loss functions. However, no general theoretical
guarantees seem to exist for this procedure, and it has not yet been extended to
the situation with competing risks.

@@

In a competing risk framework \citep{andersen2012statistical}, let
\( T\) be a time to event variable, \(D\in\{1,2\}\) the cause of the
event, and $X \in \mathcal{X}$ a vector of baseline covariates taking
values in a bounded subset \( \mathcal{X} \subset \R^p \),
\( p\in\N \). Let $\tau< \infty$ be the prediction horizon. We use
\( \mathcal{Q} \) to denote the collection of all probability measures
on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \) such that
\( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard
functions are defined by
\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \)
such that
\begin{equation*}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*}
For ease of presentation we assume throughout that the map
\( t\mapsto \Lambda_j(t \mid x) \) is continuous for all \( x \) and
\( j \). This is not a limitation: All arguments carry over directly
to the general case. We denote by \(S\) the conditional event-free
survival function:
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterised by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)=& \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=1\}}}\\
                         &  \left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}_{\tau}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

@@

As input to the super learner we need a data set
\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in \mathcal{P} \) and a collection of candidate
learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}_{\tau}\) which takes a data
set as input and returns an estimate $a(\data_n) \in \mathcal{M}_{\tau}$ of
$\Lambda_{j}$.
In what follows, we use the short-hand notation
\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
performance of \(a \in \mathcal{A}\) with a loss function
\(L\colon \mathcal{M}_{\tau} \times \mathcal{O} \rightarrow \R_+\) and
estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
cross-validation. Specifically, the expected loss of $a\in\mathcal A$
is estimated by splitting the data set $\data_n$ into $K$ disjoint
approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training sample,
while \(\data_n^{k}\) is referred to as the \(k\)'th test or hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
The oracle learner is defined as the learner that minimises the
expected loss under the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner
depend on the library of learners and on the number of folds \(K\),
and that the oracle learner is a function of the data and the unknown
data-generating distribution. However, these dependencies are
suppressed in the notation.

@@

@@

@@

@@

censoring distribution. 

@@

@@

Proposition~\ref{prop:stric-prop} can be derived from the fact that the
integrated Brier score (also called the continuous ranked probability score) is
a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
if we minimise the average loss of the integrated Brier score, we recover the
parameters of the data-generating distribution. Specifically, this implies that
the oracle of a state learner is consistent if the library of learners contains
at least one learner that is consistent for estimation of \( F \). Recall that
the function \(F\) implicitly depends on the data-generating probability measure
\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
this dependence explicit by writing \(F_0\) for the function which is obtained
by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
\eqref{eq:transition}. In the following we let
\( \mathcal{H}_{\mathcal{P}} = \{F_P : P \in \mathcal{P}\} \) where \( F_P \) is defined as in
equation~(\ref{eq:F-def}) using the measure \( P \in \mathcal{P} \).

\begin{proposition}
  \label{prop:stric-prop}
  If \(P_0\in\mathcal{P}\) then
  \begin{equation*}
    F_0 = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P_0{[\bar{B}_\tau(F, \blank)]},
  \end{equation*}
  \( H \)-almost surely for any \( j\in \{-1,0,1,2\} \) and almost any
  \( t \in [0, \tau]\).
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:proof-proposition}.
\end{proof}

@@

We establish a finite sample oracle result for the state learner. Our
Corollary~\ref{cor:oracle-prop} is in essence a special case of
Theorem 2.3 in \citep{vaart2006oracle}.  We assume that we split the
data into equally sized folds, and for simplicity of presentation we
take \( n \) to be such that \( |\data_n^{-k}| = n/K \) with \( K \)
fixed. We will allow the number of learners to grow with \( n \) and
write
\( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
\mathcal{B}_n)\) as short-hand notation and to emphasise the
dependence on \( n \).
In the following we let the space \( \mathcal{H}_{\mathcal{P}} \) be equipped with the norm
\( \| \blank \|_{P_0} \) defined as
\begin{equation}
  \label{eq:norm}
  \| F \|_{P_0} = 
  \left\{
    \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
  \right\}^{1/2}.
\end{equation}

@@

The asymptotic distribution of \( \hat{\Psi}^0 \) is
difficult to analyse due to the cross-validated model selection step
involved in the estimation of the nuisance parameters $\Lambda_1$ and
$\Lambda_2$.
Using tools from semi-parametric efficiency theory, it is possible to
construct a so-called targeted or debiased estimator with an
asymptotic distribution which we know how to estimate
\citep{bickel1993efficient,van2011targeted,chernozhukov2018double}. A
targeted estimator is based on the efficient influence function for
the parameter $\tilde{\Psi}^0$ and relies on estimators of the
nuisance parameters $\Gamma$, \( \Lambda_1 \) and $\Lambda_2$. The
efficient influence function is a \( P \)-zero mean and square
integrable function which we denote by
\( \psi(\blank ; \Lambda_1, \Lambda_2, \Gamma) \). The name is
justified because any regular asymptotically linear estimator that has
\( \psi \) as its influence function is asymptotically efficient,
meaning that it has smallest asymptotic variance among all regular
asymptotically linear estimators \citep{bickel1993efficient}.

@@

For all super learners we use a library consisting of three learners: The
Kaplan-Meier estimator \citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox
model with main effects \citep{cox1972regression, survival-package}, and a
random survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the same
library to learn the outcome distribution and the censoring distribution. Note
that the three learners in our library of learners can be used to learn the
cumulative hazard functions of the outcome and the censoring distribution. The
latter works by training the learner on the data set \( \data_n^c \), where
\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a learner
for the cumulative hazard function of the outcome to learn the cumulative hazard
function of the censoring time, we mean that the learner is trained on
\( \data_n^c \).

We compare the state learner to two IPCW based super learners: The first super
learner, called IPCW(Cox), uses a Cox model with main effects to estimate the
censoring probabilities, while the second super learner, called IPCW(KM), uses
the Kaplan-Meier estimator to estimate the censoring probabilities. The Cox
model for the censoring distribution is correctly specified in both simulation
settings while the Kaplan Meier estimator only estimates the censoring model
correctly in the simulation setting where censoring is independent. Both IPCW
super learners are fitted using the \texttt{R}-package \texttt{riskRegression}
\citep{Gerds_Ohlendorff_Ozenne_2023}. The IPCW super learners use the integrated
Brier score up to a fixed time horizon (36 months). The marginal risk of the
event before this time horizon is \(\approx 24.6\)\%. Under the `dependent
censoring' setting the marginal censoring probability before the time horizon is
\(\approx 61.9\)\%. Under the `independent censoring' setting the marginal
censoring probability before this time horizon is \( \approx 38.7 \)\%.

@@

\typeout{get arXiv to do 4 passes: Label(s) may have changed. Rerun}
